\Title

### 2021-02-23 latex/paper/main.tex
First draft of the VNHC proof in terse mode

\begin{thm}\label{thm:simply-actuated}
    Take the Hamiltonian system ~\eqref{eqn:hamiltonian}.
    Under Assumptions \ref{assm:B-const} and \ref{assm:B-perp},
    the coordinate transformation
    \(\left(\tilde{q} = \mathbf{B}q, \tilde{p} = \mathbf{B}p\right)\)
    is a canonical transformation.
    The resulting dynamics are given by 
    \begin{gather}\label{eqn:simple-hamiltonian}
        \mathcal{H}(\tilde{q},\tilde{p}) = 
        \frac{1}{2} \tilde{p}\tpose \Minv(\tilde{q}) \tilde{p} + V(\tilde{q})
        , \\
       \begin{cases}
           \dot{\tilde{q}} = \Minv(\tilde{q})\tilde{p}
           , \\
           \dot{\tilde{p}} = -\frac{1}{2} (\Id{n} \otimes \tilde{p}\tpose)
           \nabla_{\tilde{q}} \Minv(\tilde{q}) \tilde{p}
           - \nabla_{\tilde{q}} V(\tilde{q}) + \simpleB \tau
            ,
        \end{cases} \nonumber
    \end{gather}
    where 
    \(\Minv(\tilde{q}) := \mathbf{B}D^{-1}(\mathbf{B}\tpose
    \tilde{q})\mathbf{B}\tpose\)
    and
    \(V(\tilde{q}) := P(\mathbf{B}\tpose \tilde{q})\).
\end{thm}
\begin{proof}
    Since \(\mathbf{B}\) is constant, this transformation satisfies
    \(\pdiff{\tilde{q}_i}{p_m} = \pdiff{\tilde{p}_i}{q_m} = 0\) for all 
    \(i,m \in \{1,\ldots,n\}\).
    Hence, \([\tilde{q}_i, \tilde{q}_j] = [\tilde{p}_i,\tilde{p}_j] = 0\).
    Then, since \(\mathbf{B}\) is orthogonal, 
    \((\mathbf{B}_i)\tpose (\mathbf{B}\tpose)_j = \delta_i^j\).
    Hence, the Poisson bracket between \(\tilde{p}_i\) and \(\tilde{q}_j\) is 
    \begin{align*}
        [\tilde{p}_i, \tilde{q}_j] &= \sum\limits_{m=1}^n A_{i,m}A_{j,m} 
                                   = (\mathbf{B}_i)\tpose (\mathbf{B}\tpose)_j
                                   = \delta_i^j
                                   .
    \end{align*}
    By (45.10) in \cite{landau_mechanics}, this is a canonical transformation,
    so the new Hamiltonian is
    \(\mathcal{H}(\mathbf{B}\tpose q, \mathbf{B}\tpose p)\).
    Finally, since \(\dot{\tilde{p}} = \mathbf{B}\tpose \dot{p}\), the input
    matrix for the system in \((\tilde{q},\tilde{p})\) coordinates is
    \[
        \mathbf{B}B = \simpleB
        ,
    \]
    which proves the theorem.
\end{proof}
These new coordinates are called 
\textit{simply actuated coordinates}.
The first \((n-k)\) coordinates, labelled \(q_u\), are named
the \textit{unactuated coordinates}; 
the remaining \(k\) coordinates, labelled \(q_a\), are the
\textit{actuated coordinates}.
The corresponding \((p_u, p_a)\) are the \textit{unactuated} and 
\textit{actuated momenta}, respectively.

Throughout the rest of this article, we will assume the system of inquiry is
expressed in simply actuated coordinates.

@@

@@

Definition \ref{defn:vnhc} implies that the constraint manifold \(\Gamma\)
is a \(2(n-k)\)-dimensional closed embedded submanifold of 
\(\mathcal{Q} \times \R^n\).

Define the error output \(e = h(q,p)\) 
%TODO: tie this together

\begin{thm}\label{thm:vnhc-regularity}
    A relation \(h(q,p) = 0\) for system ~\eqref{eqn:simple-hamiltonian}
    is a regular VNHC of order \(k\) if and only if \(dh_{p_a} = 0\) 
    and
    \[
        \rank{\left(dh_q \Minv(q) - 
          dh_{p_u} (\Id{n-k} \otimes p\tpose)\nabla_{q_u}\Minv(q) 
         \right)\simpleB} = k
         ,
    \]
    everywhere on the constraint manifold \(\Gamma\).
\end{thm}
\begin{proof}
    \textbf{TODO: prove this in a terse way}.
\end{proof}

@@

\begin{assm}\label{assm:inertially-actuated}
    The Hamiltonian system has an inertia matrix that does not depend on the
    unactuated coordinates:
    \[
        \nabla_{q_u}M(q) = \Zmat{n(n-k) \times n}
        .
    \]
\end{assm}

\begin{thm}\label{thm:zero-dynamics}
    Let \(\mathcal{H}\) be a mechanical system in simply actuated
    coordinates satisfying Assumption \ref{assm:inertially-actuated}. 
    Let \(h(q,p_u) = 0\) be a regular VNHC of order \(k\) with constraint
    manifold \(\Gamma\). Suppose that on \(\Gamma\) one can solve for
    \(q_a\) as a function \(q_a = f(q_u,p_u)\).
    Then the constrained dynamics are given by
    \begin{equation}\label{eqn:qpu-dynamics}
        \left.\begin{aligned}
                \dot{q}_u &= \begin{bmatrix}
                    \Id{(n-k)} & \Zmat{(n-k) \times k}
                \end{bmatrix}\Minv(q)p \\
            \dot{p}_u &= -\nabla_{q_u}V(q) \\
            \end{aligned}{}\right|_{\begin{array}{c}
                q_a = f(q_u,p_u) \\ 
                p_a = g(q_u,p_u) \\
            \end{array}}
            ,
    \end{equation}
    where
    \begin{equation}\label{eqn:g-qpu}
    \begin{aligned}
        &g(q_u,p_u) := \\
           &\left(dh_q \Minv(q) \simpleB \right)\inv 
               \\
           &\left.\left(dh_{p_u} \nabla_{q_u}V(q) - dh_q \Minv(q)
        \begin{bmatrix}
            I_{n-k} \\
            \Zmat{k \times (n-k)} \\
    \end{bmatrix} p_u\right)\right|_{qa = f(q_u,p_u)}
        .
    \end{aligned}
    \end{equation}
\end{thm}
\begin{proof}
    This is a special case of Theorem 1 in \cite{nhvc_dynamic_walking} using
    \(\theta_1 = q_u\) and \(\theta_2 = p_u\).
    The proof is ommitted.
\end{proof}

>>>>

\begin{thm}\label{thm:simply-actuated}
    Take the Hamiltonian system ~\eqref{eqn:hamiltonian}.
    Under Assumptions \ref{assm:B-const} and \ref{assm:B-perp},
    the coordinate transformation
    \(\left(\tilde{q} = \mathbf{B}q, \tilde{p} = \mathbf{B}p\right)\)
    is a canonical transformation.
    The resulting dynamics are given by 
    \begin{gather}\label{eqn:simple-hamiltonian}
        \mathcal{H}(\tilde{q},\tilde{p}) = 
        \frac{1}{2} \tilde{p}\tpose \Minv(\tilde{q}) \tilde{p} + V(\tilde{q})
        , \\
       \begin{cases}
           \dot{\tilde{q}} = \Minv(\tilde{q})\tilde{p}
           , \\
           \dot{\tilde{p}} = -\frac{1}{2} (\Id{n} \otimes \tilde{p}\tpose)
           \nabla_{\tilde{q}} \Minv(\tilde{q}) \tilde{p} \\
           \phantom{---} - \nabla_{\tilde{q}} V(\tilde{q}) + \simpleB \tau
            ,
        \end{cases} \nonumber
    \end{gather}
    where 
    \(\Minv(\tilde{q}) := \mathbf{B}D^{-1}(\mathbf{B}\tpose
    \tilde{q})\mathbf{B}\tpose\)
    and
    \(V(\tilde{q}) := P(\mathbf{B}\tpose \tilde{q})\).
\end{thm}
\begin{proof}
    Since \(\mathbf{B}\) is constant, this transformation satisfies
    \(\pdiff{\tilde{q}_i}{p_m} = \pdiff{\tilde{p}_i}{q_m} = 0\) for all 
    \(i,m \in \{1,\ldots,n\}\).
    Hence, \([\tilde{q}_i, \tilde{q}_j] = [\tilde{p}_i,\tilde{p}_j] = 0\).
    Then, since \(\mathbf{B}\) is orthogonal, 
    \((\mathbf{B}_i)\tpose (\mathbf{B}\tpose)_j = \delta_i^j\).
    Hence, the Poisson bracket between \(\tilde{p}_i\) and \(\tilde{q}_j\) is 
    \begin{align*}
        [\tilde{p}_i, \tilde{q}_j] &= \sum\limits_{m=1}^n A_{i,m}A_{j,m} 
                                   = (\mathbf{B}_i)\tpose (\mathbf{B}\tpose)_j
                                   = \delta_i^j
                                   .
    \end{align*}
    By (45.10) in \cite{landau_mechanics}, this is a canonical transformation,
    so the new Hamiltonian is
    \(\mathcal{H}(\mathbf{B}\tpose q, \mathbf{B}\tpose p)\).
    Finally, since \(\dot{\tilde{p}} = \mathbf{B}\tpose \dot{p}\), the input
    matrix for the system in \((\tilde{q},\tilde{p})\) coordinates is
    \[
        \mathbf{B}B = \simpleB
        ,
    \]
    which proves the theorem.
\end{proof}
These new coordinates are called 
\textit{simply actuated coordinates}.
The first \((n-k)\) coordinates, labelled \(q_u\), are named
the \textit{unactuated coordinates}; 
the remaining \(k\) coordinates, labelled \(q_a\), are the
\textit{actuated coordinates}.
The corresponding \((p_u, p_a)\) are the \textit{unactuated} and 
\textit{actuated momenta}, respectively.

@@

In this section, we will take the system of inquiry to be a Hamiltonian
mechanical system in simply actuated coordinates.
We wish to find a controller which constraints the system according to a
function of both the configuration \(q \in \mathcal{Q}\) and the conjugate of
momentum \(p \in \R^n\).

@@

The nonholonomic constraint \(h(q,p)\) is \textit{virtual} because it is
enforced by the controller \(\tau(q,p)\), rather than arising from the physics
of the system.
The constraint manifold \(\Gamma\) is a \(2(n-k)\)-dimensional
closed embedded submanifold of \(\mathcal{Q} \times \R^n\).
Hence, the VNHC reduces the dimensionality of the dynamics from \(2n\) to
\(2(n-k)\). 
In particular, if \(k = (n-1)\), the constraint manifold is always
2-dimensional.

We often want to stabilize the constraint within a neighbourhood of \(\Gamma\).
To see when this is possible, let us define the error output \(e = h(q,p)\).
Notice that if any component of \(e_i\) has relative degree 1, we may not be able
to stabilize \(\Gamma\) -- we can always guarantee \(e_i \to 0\), but not
necessarily \(\dot{e}_i \to 0\).
It is for this reason that we define the following special type of VNHC.

\begin{defn}
    A VNHC \(h(q,p) = 0\) of order \(k\) is \textit{regular} if the output 
    \(e = h(q,p)\) is of relative degree \(\{2,2.\ldots,2\}\) everywhere on the
    constraint manifold \(\Gamma\).
\end{defn}

The authors of
\cite{nhvc_dynamic_walking,hybrid_zero_dynamics_bipedal_nhvcs,nhvc_incline_walking}
observed that a relation which uses only the unactuated conjugate of momentum
will have vector relative degree \(\{2,\ldots,2\}\).
In fact, this is a hard requirement: any VNHC which depends on the actuated
momentum cannot be regular. 
Before providing a characterization of regularity which proves this statement,
we define the following matrices as shorthand:
\begin{align}
    \mathcal{M}(q,p) &:= (\Id{n-k} \otimes p\tpose)\nabla_{q_u}\Minv(q) 
        ,\\
    \mathcal{D}(q,p_u) &:= dh_q(q,p_u) \Minv(q) 
    .
\end{align}

\begin{thm}\label{thm:vnhc-regularity}
    A relation \(h(q,p) = 0\) for system ~\eqref{eqn:simple-hamiltonian}
    is a regular VNHC of order \(k\) if and only if \(dh_{p_a} = 0\) 
    and
    \[
        \rank{\left(\mathcal{D}(q,p_u) - dh_{p_u}\mathcal{M}(q,p)\right)\simpleB} = k
         ,
    \]
    everywhere on the constraint manifold \(\Gamma\).
\end{thm}
\begin{proof}
    Let \(e = h(q,p) \in \R^k\).
    If \(dh_{p_a} \neq \Zmat{k\times k}\) for some \((q,p) \in \Gamma\), 
    then \(\tau\) appears in \(\dot{e}\) and the VNHC is not of relative degree
    \(\{2,\ldots,2\}\). Suppose now that \(dh_{p_a} = \Zmat{k\times k}\).
    Then 
    \[
        \dot{e} = \mathcal{D}(q,p_u)p - 
        dh_{p_u}\left(\frac{1}{2}\mathcal{M}(q,p)p + \nabla_{q_u}V(q)\right)
        .
    \]
    Taking one further derivative provides
    \begin{align*}
        \ddot{e} &= (\star) - 
        dh_{p_u}\left(\frac{1}{2}\diff{}{t}\left(\mathcal{M}(q,p)p\right)\right) 
        \\
                 &+\mathcal{D}(q,p_u)\simpleB \tau
    \end{align*}
    where \((\star)\) is a continuous function of \(q\) and \(p\).
    One can show that
    \begin{align*}
        dh_{p_u}\left(\frac{1}{2}\diff{}{t}\left(\mathcal{M}(q,p)p\right)\right)
        &= (\star) +\\
        &dh_{p_u} \mathcal{M}(q,p)\simpleB \tau
        .
    \end{align*}
    Hence,
    \[
       \ddot{e} = (\star) +
       \left(\mathcal{D}(q,p_u) - dh_{p_u}\mathcal{M}(q,p)\right) \simpleB \tau
        ,
    \]
    which we write as \( \ddot{e} = E(q,p) + H(q,p)\tau\) for appropriate \(E\)
    and \(H\).
    From the definition of regularity, the VNHC \(h\) is regular 
    when \(e\) is of relative degree \(\{2,\ldots,2\}\), which is true 
    if and only if the matrix premultiplying \(\tau\) is nonsingular, and hence
    that \(H\) is invertible. This proves the theorem.
\end{proof}

@@

\begin{assm}\label{assm:inertially-actuated}
    The Hamiltonian system has an inertia matrix that does not depend on the
    unactuated coordinates, so that 
    \(\nabla_{q_u}M(q) = \Zmat{n(n-k) \times n}\).
\end{assm}

\begin{thm}\label{thm:zero-dynamics}
    Let \(\mathcal{H}\) be a mechanical system in simply actuated
    coordinates satisfying Assumption \ref{assm:inertially-actuated}. 
    Let \(h(q,p_u) = 0\) be a regular VNHC of order \(k\) with constraint
    manifold \(\Gamma\). Suppose that on \(\Gamma\) one can solve for
    \(q_a\) as a function \(q_a = f(q_u,p_u)\).
    Then the constrained dynamics are given by
    \begin{equation}\label{eqn:qpu-dynamics}
        \left.\begin{aligned}
                \dot{q}_u &= \begin{bmatrix}
                    \Id{(n-k)} & \Zmat{(n-k) \times k}
                \end{bmatrix}\Minv(q)p \\
            \dot{p}_u &= -\nabla_{q_u}V(q) \\
            \end{aligned}{}\right|_{\begin{array}{c}
                q_a = f(q_u,p_u) \\ 
                p_a = g(q_u,p_u) \\
            \end{array}}
            ,
    \end{equation}
    where
    \begin{equation}\label{eqn:g-qpu}
    \begin{aligned}
        &g(q_u,p_u) := \\
        &\left(\mathcal{D}(q,p_u) \simpleB \right)\inv 
               \\
        &\left.\left(dh_{p_u} \nabla_{q_u}V(q) - \mathcal{D}(q,p_u)
        \begin{bmatrix}
            I_{n-k} \\
            \Zmat{k \times (n-k)} \\
    \end{bmatrix} p_u\right)\right|_{qa = f(q_u,p_u)}
        .
    \end{aligned}
    \end{equation}
\end{thm}
\begin{proof}
    This is a special case of Theorem 1 in \cite{nhvc_dynamic_walking} using
    \(\theta_1 = q_u\) and \(\theta_2 = p_u\).
    The proof is ommitted.
\end{proof}
